A real-time particle filter for modeling travel times of transit vehicles through a GTFS-based road network

### 2018-10-31 anzjs2018/sections/04_implementation.tex
More edits

The application consists of a
static GTFS~schedule and network data component,
and a \rt modeling and prediction component.
We chose \verb+Rcpp+ to develop the program,
providing access to other R packages for data manipulation 
(notably \verb+RSQLite+ and \verb+dplyr+),
as well as the speed and memory management capabilities of \verb|C++|. 
The following has been implemented in the R package
\verb+transitr+, available on Github (\url{https://github.com/tmelliott/transitr}).
In this section, we discuss the features of the \rt component.

The general structure of the \rt component is shown below,
with the bold steps being those discussed in this paper.
\begin{enumerate}
\item Load GTFS data from database
\item Indefinitely repeat when new data recieved \ldots
\begin{enumerate}
    \item Update or create new vehicle objects from new data
    \item \textbf{Run particle filter on each vehicle to update state}
    \item \textbf{Update state of any roads for which vehicles 
        have completed travel}
    \item Generate ETAs for vehicles uing combination of particle filter and network state
    \item Write ETAs to extended Google Protobuf binary file for distribution
\end{enumerate}
\end{enumerate}

During the development of the application,
we were primarly concerned with ensuring each component of the program
is as efficient as possible,
allowing ETAs to be generated and distributed fast enough to be faesible in real-time,
with a target of 30~seconds or faster at peak time.
Using C++ provides enough control over memory management
to ensure the program runs fast enough in \rt 
while still implementing the models.
We also use OpenMP for parallelisation,
allowing us to process multiple vehicles simultaneously,
which are each independent so threadsafety is a non-issue.

The particle filter requires thousands of particles per vehicle,
so it is necessary to explore the performance of the application
with varying number of particles.
We also assess the performance of the particle filter
for a range of fixed model parameters values,
such as system noise and GPS error.
To do so, we implemented a simulated \rt environment
in which we could analyse the same subset of real vehicle data from 8~October, 2018
using a range of settings.
These simulations were carried out of a Virtual Machine 
with 8~cores and 32~Gb of memory, 
running Ubuntu~16.04 and using R~3.4.1.

% \begin{figure}[tb]
%     \centering
%     % \includegraphics[width=0.7\textwidth]{figures/04_model_results_gps.pdf}
%     \caption{The distance between the observation and the nearest point on the shape.}
%     \label{fig:gps_dist}
% \end{figure}

@@

During development of our application,
our primary concern was the computational faesibility of a particle filter.
For each iteration, 
the timings of the individual components are recorded.
Since the number of vehicles traveling at any given time changes throughout the day,
we used the average timings over a 15~minute window to compare.

Figure~\ref{fig:timings} shows the timing results for varying number of particles.
We see that there is a very slight quadratic trend to run times,
as several steps are not of complexity $O(N)$,
the main one being the estimation of ETA quantiles.
While not discussed in this paper,
the application generates ETAs for each particle
and calculates ETA quantiles for each vehicle to demonstrate the computational faesibility.
This step requires sorting a vector, which is $O(N\log(N))$.
Clearly, our initial target of 30~seconds is obtainable given the number of particles
is not too large, and there are enough cores.

@@

The model was evaluated using several statistics,
each calculated using a range of parameter values.
The first is effective sample size,
defined in (\ref{eq:neff}),
which is used to determined if resampling is required.
Another important statistic is the \emph{degeneration rate},
or the percent of iterations where the particle filter
is unable to find any plausible states given the observation,
and the vehicle is reinitialised.
Note, however, that some observations are \emph{invalid trips},
in which the vehicle is logged in to the wrong trip,
or ``off duty'' while still transmitting for a trip.
Some of these can be filtered for being excessively far from the route,
but others---for example traveling the wrong direction---cannot be.
The last statistic we compute is the variance of road segment travel times
over time.
For this example, we varied the number of particles $N$,
GPS error $\epsilon$, and system noise $\sigma^2$.

The one parameter we can investigate from the data is GPS error, $\epsilon^2$,
by looking at the distance between observations and the shape.
Figure~\ref{fig:dist_to_route} shows the distribution of distance between the vehicle
observation and the \emph{nearest point on the path}.
We see two peaks,
one at about 1m, and another at 2.5m.
The first of these is possibly an artefact of the data,
for example automated waypoints at intersections.
We have explored a range of GPS error values from 1 to 5~m.

@@

For the effective sample size, 
increasing GPS error and system error both tend to
increase $N_\text{eff}$, as shown in Figure~\ref{fig:degen_rate}.
Increasing the GPS error distributes the weights more evenly,
while increasing system noise allows more particles to reach
the necessary speed,
so these are as expected.

@@

The degeneration rate, 
shown on the $y$-axis of Figure~\ref{fig:degen_rate},
decreases as GPS error increases,
as anomolous observations where the bus is far from the route
are more likely.
It also decreases with the number of particles,
while system noise has no apparent effect.

Lastly is the uncertainty of travel time estimates,
as shown in Figure~\ref{fig:travel_times}.
System noise again shows no effect on the variability of travel time estimates,
however increasing GPS error tends to increase travel time variability.
This means that there is a trade-off between the models performance
and how well it estimates the main parameter of interest:
larger values of $\epsilon$ result in less degeneration and larger $N_\text{eff}$,
at the cost of more variable travel time estimates.

>>>>

The application consists of two components:
the first handles importing static GTFS data and constructing the network,
while the second implements the \rt models and prediction.
We chose \verb+Rcpp+ to develop the program,
which provides access to other R packages for data manipulation 
(notably \verb+RSQLite+ and \verb+dplyr+),
as well as the speed and memory management capabilities of \verb|C++|. 
The program is implemented in the R package
\verb+transitr+, available on Github (\url{https://github.com/tmelliott/transitr}).
In this section, we discuss the features of the \rt component
and assess its performance in \rt,
both with respect to timing and parameter estimation.

The general structure of the \rt component is shown below,
with the bold steps being those discussed in this paper.
\begin{enumerate}
\item Load GTFS data from database
\item Each time new data are recieved \ldots
\begin{enumerate}
    \item Update or create new vehicle objects from the new data
    \item \textbf{Run particle filter on each vehicle to update or initialise state}
    \item \textbf{Update state of any roads for which vehicles 
        have completed travel}
    \item Generate ETAs for vehicles uing combination of particle filter and network state
    \item Write ETAs to extended Google Protobuf binary file for distribution
\end{enumerate}
\end{enumerate}

During the development of the application,
we were primarly concerned with ensuring each component of the program
is as efficient as possible,
allowing ETAs to be generated and distributed fast enough to be faesible in real-time,
with a target of 30~seconds or faster at peak time.
Using C++ provides the memory management control necessary to make the application
viable in real-time.
We also make use of OpenMP for parallisation,
since each vehicle is modelled independently so it is trivial to scale up without
worrying about threadsafety.

The number of particles needed per vehicle 
depends on many factors,
so it is necessary to explore the performance of the application
with varying number of particles.
Application performance is also assessed
for a range of fixed model parameter values,
such as system noise and GPS error.
To enable comparisons, we implemented a simulated \rt environment
in which the same subset of real vehicle data from 8~October, 2018
could be analysed using a range of settings.
These simulations were carried out of a Virtual Machine 
with 8~cores and 32~Gb of memory, 
running Ubuntu~16.04 and using R~3.4.1.

@@

In each iteration, 
the timing of the various program components is recorded.
Since the number of vehicles traveling at any given time changes throughout the day,
we used the average timings over an off-peak 15~minute window.
Figure~\ref{fig:timings} shows the average timings for 
varying numbers of particles, $N$.

The most time consuming component is ETA writing,
which involves summarising the individual ETAs estimated for each particle.
While not discussed here this involves quantiles which use a sorting algorithm,
which has complexity $O(N \log N)$ 
which explains the slightly quadratic trend in the timings.

The next most intensive step is vehicle updating,
and takes about 5~seconds for 8000 particles.
The ETA prediction step only takes a few seconds,
although this may increase once a more comprehensive model is developed.
On our 8-core virtual machine the entire process is completed within 20~seconds
when 8000~particles are used.
In the next section we explore the effect of changing $N$
to allow finding the optimal number.

@@

For each simulation, 
the model was evaluated using several statistics.
These were effective sample size, $N_\text{eff}$ calculated using (\ref{eq:neff});
degeneration rate, the percentage of iterations in which the vehicle was lost
(no plausible particles);
and the variance of segment travel time estimates.

The GPS error parameter values were chosen by examining the distribution
of the distance between observations and the nearest point on the route.
This is shown in Figure~\ref{fig:dist_to_route},
in which we see a peak at about 1m and another at 2.5m,
so we have chosen GPS error values of $\epsilon \in \{1,2,3,5\}$.
System error is the variation of speed per second,
or acceleration rate,
so our chosen values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$ 
correspond to a variance of 0.09, 0.9, 9, and 45 meters per second over 
30~seconds.

@@

From Figure~\ref{fig:degen_rate}, we see that $N_\text{eff}$ increases
with GPS error and decreases with system noise.
The degeneration rate decreases with GPS error and increases with $N$,
but is relatively unaffected by system noise.
Large GPS error affects the likelihood,
and gives more weight to particles farther from the observation
than does a smaller GPS error.
Conversesly, increasing system noise spreads out the particle cloud,
so fewer particles will be near the vehicle.
Thus, these results are not surprising, 
but show that the model is working as expected.

@@

The last measure of performance is the variance of segment travel times.
This is calculated by \_\_\_.
Figure~\ref{fig:travel_times} shows that variance increases with GPS error,
and decreases with $N$,
with little effect made by system noise.
There is therefore a tradeoff between particle filter performance
($N_\text{eff}$ and degeneration rate) and parameter estimation,
although the true test will be calculating the prediction error
of ETAs once the prediction model has been implemented.

### 2018-10-31 anzjs2018/sections/06_discuss.tex
More edits

In this paper we have described an approach to transit vehicle modeling
that enables the \rt estimation of road state from vehicle position data,
which will in future be used to make what will hopefully be more reliable 
arrival time predictions.
The vehicle model, implemented using a particle filter,
allows complex bus behaviours to be modeled,
with future work planned to explore the other parameters---dwell time and stopping probabilities---%
as well as to extend the model to also account for intersections.
The travel times estimated from the particle filter then allow us to model \rt road state.

The network model presented here is of the simplest nature,
but provides the foundation for future work which will involve
deveoping a formal model that combines \rt information with
historical data, allowing predictions of future road state to be more accurate,
for example incorporating peak time congestion forecasts.
This in turn provides the basis for an arrival time prediction model,
which will be a combination of vehicle state and road state forecasts.

>>>>

In this paper we have described an approach to transit vehicle modeling
that enables the \rt estimation of road state from vehicle position data,
which will in future be used to make what will hopefully be more reliable 
arrival time predictions.
The vehicle model, implemented using a particle filter,
allows complex bus behaviours to be modeled,
with future work planned to explore the other parameters---dwell time and stopping probabilities---%
as well as to extend the model to also account for physical intersections.
The travel times estimated from the particle filter then allow modeling of
\rt road state.

The network model presented here is simple,
but provides the foundation for future work which will involve
developing a formal model that combines \rt information with
historical data, allowing for more accurate predictions of future road state,
for example incorporating peak time congestion forecasts.
This in turn will provide the basis for an arrival time prediction model,
a combination of vehicle state and road state forecasts.
